---
title: "Installing the Jmisc, Jutils, PEB, SIM and other R packages"
author: "Jason Thorpe [ jdthorpe_at_gmail.com ]"
date: "2014-12-25"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Jmisc}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```
## Installing R and Required R packages

The latest release for R can be found at:

	[http://cran.fhcrc.org/](http://cran.fhcrc.org/)

by clicking on the link for "R Binaries" at the left side of the page, 
following the link for your computer’s OS (windows, most likely), and then
by clicking on the link to 'base' and then click "Download R".

Once R is installed, you will need to install some pacagges as follows. 

```{r, eval = FALSE}
# intall the package builder  packages
install.packages(c('roxygen2','devtools'),repos="http://cran.fhcrc.org")

# intall the required dependencies 
install.packages(c('CompQuadForm','multilevel'),repos="http://cran.fhcrc.org")

# intall the optional dependencies 
install.packages(c('RODBC'),repos="http://cran.fhcrc.org")

# intall the vignette builder packages (optional)
install.packages(c('pander'),repos="http://cran.fhcrc.org")
```

You can also install these packages using the 'packages' menu, if you prefer.

## Installing Packages the Easy Way 

### Option 1 (Windows Only)

Local packages can be installed from the binary '.zip' files either 
(a) by selecting "Install package(s) from local zip files..." from the 
"packages menu" or (b) from the command line via:

```{r, eval = FALSE}
home <- "R:/Urban_N/UrbanGrp/Data Analysis/JThorpe/RTools/pkg-bin"
zipFiles <- c("Jmisc_0.1.zip",
				   "Jutils_0.1.zip",
				   "mfactor_0.1.zip",
				   "PEB_0.1.zip",
				   "sim_0.1.zip")
install.packages(file.path(home,zipFiles), repos = NULL)
```

### Option 2 (OSX Only )

To intall source packages using the OSX GUI, choose "Package Installer" from the 
"Packages & Data" drop down menu, and at the top of the "R Package Installer" window,
choose "Local Source Package"  from the drop down menu at the top (which by
default is set to "CRAN (binaries)") and then click "install".   This will bring
up a file menu, and you just need to select the approprioate '.tar.gz' file.

### Option 3 (Any Platform, Including OSX and Windows)

Local packages can be installed from source packages at the command line via:

```{r , eval = FALSE}
home <- "R:/Urban_N/UrbanGrp/Data Analysis/JThorpe/RTools/pkg-src"
tarFiles <- c("Jmisc_0.1.tar.gz",
				   "Jutils_0.1.tar.gz",
				   "mfactor_0.1.tar.gz",
				   "PEB_0.1.tar.gz",
				   "sim_0.1.tar.gz")
install.packages(file.path(home,tarFiles), repos = NULL,type='source')
```


## Installing Local Packages the Hard Way 

On my Windows machine at the Hutch, my library path points to 

```
.libPaths()
#> [1] "\\\\home.fhcrc.org/jthorpe/R/win-library/3.1"
#> [2] "C:/Program Files/R/R-3.1.0/library"
```

which is a problem devtools::install fails when attempting to install a
package onto a server location (e.g.`"\\home.fhcrc.org/jthorpe/...`').  
These lines fixt that bug:

```{r}
.libPaths(gsub('^\\\\\\\\home.fhcrc.org(\\\\|/)[-a-zA-Z0-9._]+(\\\\|/)?','h:\\\\',.libPaths()))

root <- "R:/Urban_N/UrbanGrp/Data Analysis/JThorpe/RTools"
```

Alternative

Before installing, you will want to document any changed packages as
follows, and after installing, it is a good idea to re-build the vignettes.


```{r, eval = FALSE}

devtools::document(file.path(root,"Jmisc"))
devtools::install(file.path(root,"Jmisc"))
devtools::build_vignettes(file.path(root,"Jmisc"))

devtools::document(file.path(root,"Jutils"))
devtools::install(file.path(root,"Jutils"))

devtools::document(file.path(root,"mfactor"))
devtools::install(file.path(root,"mfactor"))
devtools::build_vignettes(file.path(root,"mfactor"))

devtools::document(file.path(root,"PEB"))
devtools::install(file.path(root,"PEB"))
devtools::build_vignettes(file.path(root,"PEB"))

devtools::document(file.path(root,"SIM"))
devtools::install(file.path(root,"SIM"))
devtools::build_vignettes(file.path(root,"SIM"))

devtools::document("H:/O/popsim/PopSim/dist/LifeTable")
devtools::install("H:/O/popsim/PopSim/dist/LifeTable")
devtools::build_vignettes("H:/O/popsim/PopSim/dist/LifeTable")
```
Note that the package SIM depends on the package `mfactor`, so you may need to
have `mfactor` installed befor you try and install `sim`


## Using Local Packages 

Once installed, these packages can be loaded in the usual way 

```{r, eval = FALSE}
require('Jmisc')
require('Jutils')
require('PEB')
require('mfactor')
require('sim')
```

Package contents can be explored in the usual way. 
```{r, eval = FALSE}
help(package='Jmisc')
help(package='Jutils')
help(package='PEB')
help(package='mfactor')
help(package='sim')
```

and help files are available for all exported functions!

```{r, eval = FALSE}
?mfactor
?ppeb
?lu
```

# Building Packages

This section is concerned with building R packages, and is not required
for using existing local package files.

## Building Vignettes

### Step 0.1: Install Pandoc + Bug Fix
Building Vignettest in the the PEB and Jmisc packages required that you install
PanDoc. For details on installing Pandoc click [here](http://johnmacfarlane.net/pandoc/installing.html).

There is a bug in the windows version of devtools, which prevents it from
finding Pandoc, which will result in an error message stating `Error :
invalid version specification ‘pandoc.exe 1.13.0.1’`.  As a work
around, you can set an environment variable similar to:

```{r, eval = FALSE}
Sys.setenv(RSTUDIO_PANDOC="C:\\Users\\jthorpe\\AppData\\Local\\Pandoc\\")
```

For details on the bug fix, see [this article](http://stackoverflow.com/questions/26803652/devtoolsbuild-vignettes-yields-error-invalid-version-specification-pandoc) on StackOverFlow.

### Step 0.2: Un-Fix the Bug Fix
Remember that bug in `devtools::install` that was fixed by resetting the
library paths?  Well that fix is a bug here because PanDoc can't see
certian drive mappings (e.g. `"h:\..."`). So if you installed any of the 
packages using `devtools::install` you'll need to re-start your session
so that .libPaths() does not rely on a drive mapping that PanDoc can't
follow.

### Step 1: Build the Vignettes

```{r, eval = FALSE}
devtools::build_vignettes(file.path(root,"Jmisc"))
devtools::build_vignettes(file.path(root,"PEB"))
devtools::build_vignettes(file.path(root,"SIM"))
devtools::build_vignettes(file.path(root,"mfactor"))
devtools::build_vignettes("H:/O/popsim/PopSim/dist/LifeTable")


```


## Building Package Files

To build the source packages (including the vignettes), use:

```{r, eval = FALSE}
devtools::build(file.path(root,"Jmisc")		,path=file.path(root,'pkg-src'))
devtools::build(file.path(root,"Jutils")	,path=file.path(root,'pkg-src'))
devtools::build(file.path(root,"PEB")		,path=file.path(root,'pkg-src'))
devtools::build(file.path(root,"mfactor")	,path=file.path(root,'pkg-src'))
devtools::build(file.path(root,"SIM")		,path=file.path(root,'pkg-src'))
```

and to build the (windows) binary packages (including the vignettes), use:

```{r, eval = FALSE}
devtools::build(file.path(root,"Jmisc")		,path=file.path(root,'pkg-bin'),binary=TRUE)
devtools::build(file.path(root,"Jutils")	,path=file.path(root,'pkg-bin'),binary=TRUE)
devtools::build(file.path(root,"PEB")		,path=file.path(root,'pkg-bin'),binary=TRUE)
devtools::build(file.path(root,"mfactor")	,path=file.path(root,'pkg-bin'),binary=TRUE)
devtools::build(file.path(root,"SIM")		,path=file.path(root,'pkg-bin'),binary=TRUE)
```

## Final Notes

These pacakges were initialized via:

```{r, eval = FALSE}
dir <- "R:/Urban_N/UrbanGrp/Data Analysis/JThorpe/RTools/Jmisc"
devtools::create(dir)
devtools::use_vignette('Jmisc-readme',dir)
```

And I added this line each to the first code block in each 
vignette to make the code code output pretty

	```{r, eval = FALSE}
	knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
	```


## Windows Bugs 

### Bug 1: Libraryies on a remote machine.

On my Windows machine at the Hutch, my library path points to 

```
.libPaths()
#> [1] "\\\\server.name/me/R/win-library/3.1"
#> [2] "C:/Program Files/R/R-3.1.0/library"
```

which is a problem because devtools::install fails when attempting to install a
package onto a server location (e.g.`"\\server.name/me/...`').  

There are two methods for fixing this bug.  The first (and preferred) method is to 
set an enviromment variable by choosing Start Menu > Control Panel > System > 
Advanced system settings > Environment variables, and click "new" to create a new 
system varible called `R_LIB_USER` which points to a directory on your local machine 
where you would like your R User Library to reside (e.g. `C:\Users\Me\R\win-library\3.1`)

The second solution is to map the server to a drive letter and update your `.libPaths()`
accorginly like so:

```{r}
.libPaths(gsub('^\\\\\\\\server.name.fhcrc.org(\\\\|/)[-a-zA-Z0-9._]+(\\\\|/)?',
               'h:\\\\',.libPaths())) # this presumes the network drive is mapped to 'h'
```

Note that this solution requires that you either put the above call in your `.Rprofile` or 
you call the line before installing the package. 

Note that this is not preferred b/c this alternate fix actually introduces a bug in Pandoc
which is used to build the vignettes...

### Bug 1: Pandoc and RStudio

Pandoc expects that Rstudio is installed, and if it isn't, it freaks out.  The preferred 
(permenaent fix) is to create an environment variable called `RSTUDIO_PANDOC` with the value 
`C:\Users\jthorpe\AppData\Local\Pandoc` (Bug 1 for instructions for creating an environment varaible).

Alternatively, you can call this line prior to building the vignettes.

	Sys.setenv(RSTUDIO_PANDOC="C:\\Users\\jthorpe\\AppData\\Local\\Pandoc\\")

This fix is not the preferred method because it does not work with `R CMD check` 
(and hence `devtools::check()`) which rely on a `--Vanilla` session, which won't 
include the above call to `Sys.setenv()`




